Abstract

### 2025-10-24 output/Main_text.Rmd
Correct interpretation of parameters
Stating it may affect the test scoring was incorrect; now it refers to the information an item provides about each trait and its quality assessing it.

We have seen that taking the latent space structure into account may have a
substantial effect on the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Interpreting the item properties correctly requires a solid understanding of
their relationship with the latent covariance structure,
and we have shown that this can only be achieved
with a proper geometrical representation.
Simple items are not affected much by this issue but,
as MIRT evolves and converges with the common factor psychometric tradition,
we find MIRT models applied more often to new problems,
involving complex multidimensional item structures
(even including items with opposite-signed loadings).
Our results will be especially relevant in those cases,
where the interpretation of the items may be affected more drastically.
For example, the relative importance of an item in scoring a given trait
could be downplayed if it is strongly aligned with that trait dimension
but this alignment is overlooked.
Once again, in multi-group applications,
where differences in the latent space structure are critical,
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Different covariance structures across groups may significantly affect them
due to differences in either the variances or the correlation structure.
Therefore, even when the model parameters are equal across groups,
the multidimensional parameters may not be the same
if the covariance structures differ.
Conversely, the same multidimensional parameters may correspond
to different parameter models.
The subsequent invariance of either the M2PL parameters
or the multidimensional ones may lead to different interpretations of
the same model when applied to different groups.
Although this is an empirical question that requires further research,
it could have significant implications for future theoretical developments
in the understanding of MIRT theory.

>>>>

We have seen that taking the latent space structure into account may have a
substantial effect on the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Interpreting the item properties correctly requires a solid understanding of
their relationship with the latent covariance structure,
and we have shown that this can only be achieved
with a proper geometrical representation.
Simple items are not affected much by this issue but,
as MIRT evolves and converges with the common factor psychometric tradition,
we find MIRT models applied more often to new problems,
involving complex multidimensional item structures
(even including items with opposite-signed loadings).
Our results will be especially relevant in those cases,
where the interpretation of the items may be affected more drastically.
For example, using the agnostic version of the parameters and/or an orthogonal plot,
the alignment of an item with the trait axes could be easily misinterpreted.
Then, the information it provides about each of those traits
(and, subsequently, its quality for assessing them)
could be either downplayed or exaggerated.
Once again, this is paramount in multi-group applications,
where differences in the latent space structure are critical and
the agnostic version could overlook relevant differences in
the values of the multidimensional parameters.
Different covariance structures across groups may significantly affect them
due to differences in either the variances or the correlation structure.
Therefore, even when the model parameters are equal across groups,
the multidimensional parameters may not be the same
if the covariance structures differ.
Conversely, the same multidimensional parameters may correspond
to different parameter models.
The subsequent invariance of either the M2PL parameters
or the multidimensional ones may lead to different interpretations of
the same model when applied to different groups.
Although this is an empirical question that requires further research,
it could have significant implications for future theoretical developments
in the understanding of MIRT theory.
